---
title: "Mini-Project #04: Exploring Recent US Political Shifts"
author: Chenbin Wu
output: html
---

## Introduction

This report presents an analysis of the county-level results from the 2020 and 2024 U.S. presidential elections, focusing on the counties that showed significant Democratic support. Although the national election results were close, examining the geographic patterns reveals how Democratic candidates, particularly Joe Biden, gained significant traction in key regions across the U.S. This report highlights the Democratic strongholds that emerged or grew in 2024 and demonstrates the continued resilience of Democratic voters in certain areas. Through this analysis, we can better understand how and why the Democratic Party has seen growth and how these trends might influence future elections.

## Data Collection and Preperation

The first step in this analysis was gathering U.S. county boundary data from the Census Bureau. This shapefile data was downloaded and loaded into R to map county-level election results. This data was crucial in providing a visual representation of the geographical distribution of votes.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
# Load Libraries
library(sf)
library(httr2)
library(rvest)
library(dplyr)
library(janitor)
library(stringr)
library(readr)
library(tidyr)
library(ggplot2)

# Setup Directories and File Paths
dir_path <- "data/mp04"
zip_url <- "https://www2.census.gov/geo/tiger/GENZ2023/shp/cb_2023_us_county_500k.zip"
zip_file <- file.path(dir_path, "cb_2023_us_county_500k.zip")

# Create directories and download shapefile if necessary
if (!dir.exists(dir_path)) {
  dir.create(dir_path, recursive = TRUE)
  message("Created directory: ", dir_path)
}

if (!file.exists(zip_file)) {
  download.file(zip_url, destfile = zip_file, mode = "wb")
  message("Downloaded shapefile to: ", zip_file)
} else {
  message("Shapefile already downloaded.")
}

# Unzip shapefile if not already extracted
unzip_dir <- file.path(dir_path, "cb_2023_us_county_500k")
shp_file <- file.path(unzip_dir, "cb_2023_us_county_500k.shp")

if (!file.exists(shp_file)) {
  unzip(zip_file, exdir = unzip_dir)
  message("Unzipped shapefile to: ", unzip_dir)
} else {
  message("Shapefile already unzipped.")
}

# Load shapefile
counties_sf <- st_read(shp_file)
glimpse(counties_sf)
```

### Collecting 2024 Data

For the 2024 election, county-level results were gathered through web scraping from each state’s Wikipedia page. This process ensured that the data was up-to-date and comprehensive. The results were then cleaned and organized by state for further analysis. Some states, like Washington, were missing results, but these gaps were identified for review.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
counties_sf <- st_transform(counties_sf, crs = 4326)

# Scraper function for 2024 results
get_state_results <- function(state_name) {
  state_slug <- gsub(" ", "_", state_name)
  wiki_url <- paste0("https://en.wikipedia.org/wiki/2024_United_States_presidential_election_in_", state_slug)
  
  dir_path <- "data/mp04/html_pages"
  if (!dir.exists(dir_path)) dir.create(dir_path, recursive = TRUE)
  
  html_file <- file.path(dir_path, paste0(state_slug, ".html"))
  if (!file.exists(html_file)) {
    req <- request(wiki_url) |> req_perform()
    writeBin(resp_body_raw(req), html_file)
    message("Downloaded and saved HTML for: ", state_name)
  } else {
    message("Using cached HTML for: ", state_name)
  }
  
  page <- read_html(html_file)
  tables <- page |> html_nodes("table") |> html_table(fill = TRUE)
  
  county_keywords <- c("County", "Parish", "Borough", "District")
  target_table <- NULL
  
  # Find the county-level table
  for (tbl in tables) {
    if (any(str_detect(names(tbl), paste(county_keywords, collapse = "|")))) {
      target_table <- tbl
      break
    }
  }
  
  if (is.null(target_table)) {
    warning("No suitable county-level table found for: ", state_name)
    return(NULL)
  }
  
  target_table <- target_table |> clean_names() |> mutate(state = state_name)
  return(target_table)
}

# Scrape all states for 2024 election results
state_names <- state.name
all_results_2024 <- lapply(state_names, get_state_results)
all_results_2024_df <- bind_rows(Filter(Negate(is.null), all_results_2024))

# Logging missing data
missing_states_2024 <- state_names[sapply(all_results_2024, is.null)]
print(missing_states_2024)
```

### Collecting 2020 Data

Similarly, the 2020 election data was scraped and cleaned in a similar manner to the 2024 data. Any missing results were identified, and the data was prepared for further comparison and analysis.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
# Scraper function for 2020 results (similar to 2024, but with election year set to 2020)
get_state_results_2020 <- function(state_name) {
  state_slug <- gsub(" ", "_", state_name)
  wiki_url <- paste0("https://en.wikipedia.org/wiki/2020_United_States_presidential_election_in_", state_slug)
  
  dir_path <- "data/mp04/html_pages_2020"
  if (!dir.exists(dir_path)) dir.create(dir_path, recursive = TRUE)
  
  html_file <- file.path(dir_path, paste0(state_slug, ".html"))
  if (!file.exists(html_file)) {
    req <- request(wiki_url) |> req_perform()
    writeBin(resp_body_raw(req), html_file)
    message("Downloaded and saved HTML for 2020 election: ", state_name)
  } else {
    message("Using cached HTML for 2020 election: ", state_name)
  }
  
  page <- read_html(html_file)
  tables <- page |> html_nodes("table") |> html_table(fill = TRUE)
  
  county_keywords <- c("County", "Parish", "Borough", "District")
  target_table <- NULL
  
  # Find the county-level table for 2020 results
  for (tbl in tables) {
    if (any(str_detect(names(tbl), paste(county_keywords, collapse = "|")))) {
      target_table <- tbl
      break
    }
  }
  
  if (is.null(target_table)) {
    warning("No suitable county-level table found for 2020 election: ", state_name)
    return(NULL)
  }
  
  target_table <- target_table |> clean_names() |> mutate(state = state_name, election_year = 2020)
  return(target_table)
}

# Scrape all states for 2020 election results
all_results_2020 <- lapply(state_names, get_state_results_2020)
all_results_2020_df <- bind_rows(Filter(Negate(is.null), all_results_2020))

# Logging missing data for 2020 results
missing_states_2020 <- state_names[sapply(all_results_2020, is.null)]
print(missing_states_2020)
```

### Cleaning and Aggregating Election Data for 2020 and 2024

Once both the 2020 and 2024 election data were collected, the next step was to clean and standardize the county-level results. This included handling missing values and ensuring consistency across counties and states, particularly important for identifying where Democratic candidates performed best.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
all_results_2020_df <- all_results_2020_df |>
  mutate(county = str_to_title(county))  # Normalize county names for better matching
all_results_2024_df <- all_results_2024_df |>
  mutate(county = str_to_title(county))

# Clean and standardize county names for 2020
clean_2020 <- all_results_2020_df |>
  filter(
    !county %in% c("County", "Total", "Parish", "Borough", "District"),
    !str_detect(county, "Total|County|Parish|Borough|District")
  ) |>
  mutate(
    county = str_replace_all(county, " County| Parish| Borough| District", ""),
    county = str_to_title(str_trim(county))
  )

# Convert vote counts to numeric for 2020
clean_2020 <- clean_2020 |>
  mutate(across(
    c(donald_trump_republican, joe_biden_democratic),
    ~ as.numeric(gsub(",", "", .)),
    .names = "{.col}_num"
  ))

# Aggregate by county and state for 2020
agg_2020 <- clean_2020 |>
  group_by(state, county) |>
  summarise(
    trump_votes = sum(donald_trump_republican_num, na.rm = TRUE),
    biden_votes = sum(joe_biden_democratic_num, na.rm = TRUE),
    .groups = "drop"
  )

counties_2020_sf <- counties_sf |>
  mutate(
    NAME = str_to_title(NAME),
    STATE_NAME = str_to_title(STATE_NAME)  # Ensure this matches the actual column name
  ) |>
  left_join(agg_2020, by = c("NAME" = "county", "STATE_NAME" = "state"))

# Clean and standardize county names for 2024
clean_2024 <- all_results_2024_df |>
  filter(
    !county %in% c("County", "Total", "Parish", "Borough", "District"),
    !str_detect(county, "Total|County|Parish|Borough|District")
  ) |>
  mutate(
    county = str_replace_all(county, " County| Parish| Borough| District", ""),
    county = str_to_title(str_trim(county))
  )

# Convert vote counts to numeric for 2024
clean_2024 <- clean_2024 |>
  mutate(across(
    c(donald_trump_republican, kamala_harris_democratic),
    ~ as.numeric(gsub(",", "", .)),
    .names = "{.col}_num"
  ))

# Aggregate by county and state for 2024
agg_2024 <- clean_2024 |>
  group_by(state, county) |>
  summarise(
    trump_votes = sum(donald_trump_republican_num, na.rm = TRUE),
    harris_votes = sum(kamala_harris_democratic_num, na.rm = TRUE),
    .groups = "drop"
  )

counties_2024_sf <- counties_sf |>
  mutate(
    NAME = str_to_title(NAME),
    STATE_NAME = str_to_title(STATE_NAME)  # Ensure this matches the actual column name
  ) |>
  left_join(agg_2024, by = c("NAME" = "county", "STATE_NAME" = "state"))
```

### Initial Analyzation

1.  **Largest County to Vote for Trump in 2024:**

In the 2024 presidential election, the county with the highest number of votes for Donald Trump was identified by analyzing geographic voting data. The top-performing county was found through a filter for valid Trump vote counts, highlighting where Trump secured the largest numerical support. This county’s results underscore the regions where Trump continued to garner significant backing, even though the overall victory in a county might not have been his.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
counties_2024_sf |>
  filter(!is.na(trump_votes)) |>
  slice_max(trump_votes, n = 1) |>
  select(NAME, STATE_NAME, trump_votes)
```

2.  **County with Most Vote for Biden in 2020:**

Kalawao County in Hawaii stands out as the county with the highest percentage of votes for Joe Biden in 2020. With an overwhelming 95.83% of the vote going to Biden, this result illustrates a near-total alignment of the county’s voters with the Democratic ticket. This extreme concentration of support reflects Kalawao’s strong Democratic lean and indicates how some regions have solidified their political allegiance, possibly due to local demographics or longstanding political traditions.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
counties_2020_sf |>
  filter(!is.na(trump_votes) & !is.na(biden_votes)) |>
  mutate(total_votes = trump_votes + biden_votes,
         biden_share = biden_votes / total_votes) |>
  slice_max(biden_share, n = 1) |>
  select(NAME, STATE_NAME, biden_votes, total_votes, biden_share)
```

3.  **Counties with the Largest Shift in Vote for Trump:**

A key finding is that Gwinnett County, Georgia, saw the most significant increase in Trump’s votes from 2020 to 2024, with an impressive surge of 173,041 votes. This substantial uptick suggests a shift in the county's political landscape, potentially driven by demographic or ideological changes in the electorate. In contrast, Kalawao County in Hawaii, where Biden's support was overwhelmingly high in 2020, remained a Democratic stronghold. The differences between these counties highlight the complex, shifting political dynamics across the U.S. and are indicative of broader national trends.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
vote_change <- counties_2020_sf |>
  st_drop_geometry() |>
  group_by(NAME, STATE_NAME) |>
  summarise(trump_votes_2020 = sum(trump_votes, na.rm = TRUE), .groups = "drop") |>
  inner_join(
    counties_2024_sf |>
      st_drop_geometry() |>
      group_by(NAME, STATE_NAME) |>
      summarise(trump_votes_2024 = sum(trump_votes, na.rm = TRUE), .groups = "drop"),
    by = c("NAME", "STATE_NAME")
  ) |>
  mutate(trump_shift = trump_votes_2024 - trump_votes_2020) |>
  slice_max(trump_shift, n = 1) |>
  select(NAME, STATE_NAME, trump_shift)
vote_change
```

4.  **State with the Largest Shift Towards Harris in 2024:**

Florida exhibited the most substantial shift in Trump’s vote share between the 2020 and 2024 elections. The increase in support for Trump in Florida signals a growing Republican presence in the state, potentially driven by factors such as demographic changes, voter preferences, or political polarization. This shift may be indicative of broader national trends and highlights the importance of battleground states in future elections, with Florida emerging as a critical state in the 2024 election.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
state_shift <- counties_2020_sf |>
  st_drop_geometry() |>
  group_by(STATE_NAME) |>
  summarise(
    trump_2020 = sum(trump_votes, na.rm = TRUE),
    biden_2020 = sum(biden_votes, na.rm = TRUE),
    .groups = "drop"
  ) |>
  inner_join(
    counties_2024_sf |>
      st_drop_geometry() |>
      group_by(STATE_NAME) |>
      summarise(
        trump_2024 = sum(trump_votes, na.rm = TRUE),
        harris_2024 = sum(harris_votes, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "STATE_NAME"
  ) |>
  mutate(trump_shift = trump_2024 - trump_2020) |>
  slice_min(trump_shift, n = 1) |>
  select(STATE_NAME, trump_2020, trump_2024, trump_shift)
state_shift
```

5.  **The Largest County, by Area, in This Dataset:**

Yukon-Koyukuk, Alaska, stands out as the largest county by area in this dataset. Its vast geographical expanse could indicate lower population density, which might influence voting patterns and political strategies. Understanding the spatial dynamics of such large regions is important for electoral campaigns and resource allocation, as larger areas often have fewer polling places or longer travel distances for voters.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
counties_sf |>
  slice_max(ALAND, n = 1) |>
  select(NAME, STATE_NAME, ALAND)

# Calculate Trump percentage for 2020
counties_2020_sf <- counties_2020_sf |>
  mutate(
    total_votes_2020 = trump_votes + biden_votes,  # Total votes for 2020
    trump_pct_2020 = trump_votes / total_votes_2020 * 100  # Trump percentage
  )

# Calculate Trump percentage for 2024
counties_2024_sf <- counties_2024_sf |>
  mutate(
    total_votes_2024 = trump_votes + harris_votes,  # Total votes for 2020
    trump_pct_2024 = trump_votes / total_votes_2024 * 100  # Trump percentage
  )
```

6.  **County with the Highest Voter Density in 2020?**

In 2020, St. Louis, Missouri, emerged as the county with the highest voter density, with 527,644 votes cast. This high density suggests a highly engaged electorate, potentially reflecting urban areas where political engagement and turnout are often stronger. Voter density can also be a proxy for the intensity of political campaigns, as counties with more concentrated populations may have more competitive races.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
counties_2020_sf |>
  mutate(total_votes = trump_votes + biden_votes,
         voter_density = total_votes / ALAND) |>
  slice_max(voter_density, n = 1) |>
  select(NAME, STATE_NAME, total_votes, ALAND, voter_density)
```

7.  **County with the Largest Increase in Voter Turnout in 2024?**

Montgomery County, Texas, saw the largest increase in voter turnout from 2020 to 2024, with a remarkable jump of 36,482 votes. This surge indicates growing political engagement, which could reflect a shift in local dynamics or an increased focus on voter outreach and participation efforts. The significant increase in voter turnout could be a key factor in understanding future election trends, particularly in swing states or suburban areas where demographic changes are occurring.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"
turnout_change <- counties_2020_sf |>
  st_drop_geometry() |>
  mutate(total_2020 = trump_votes + biden_votes) |>
  select(NAME, STATE_NAME, total_2020) |>
  inner_join(
    counties_2024_sf |>
      st_drop_geometry() |>
      mutate(total_2024 = trump_votes + harris_votes) |>
      select(NAME, STATE_NAME, total_2024),
    by = c("NAME", "STATE_NAME")
  ) |>
  mutate(turnout_diff = total_2024 - total_2020) |>
  slice_max(turnout_diff, n = 1) |>
  select(NAME, STATE_NAME, total_2020, total_2024, turnout_diff)

print(turnout_change)
```

### County-Level Shift in Trump Vote Share: 2020 to 2024

The county-level shift in Trump’s vote share between 2020 and 2024 reveals significant political changes across the U.S. While rural areas generally show stronger support for Trump, urban and suburban counties are exhibiting slight shifts away from him, indicating evolving political preferences. This trend is especially pronounced in battleground states, signaling that the 2024 election will remain fiercely contested. These shifts, as visualized in the mapped data, suggest a further deepening of partisan divides and a more polarized political landscape.

```{r}
#| echo: true
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"

# Task 5, Step1: Computing the shift rightwards for each county
# Convert the sf objects to data frames before the join
counties_2020_df <- counties_2020_sf |>
  st_drop_geometry() |>
  select(NAME, STATE_NAME, trump_pct_2020)

counties_2024_df <- counties_2024_sf |>
  st_drop_geometry() |>
  select(NAME, STATE_NAME, trump_pct_2024)

# Perform the join on the data frames
# First, merge the 2020 and 2024 vote data
county_votes <- counties_2020_df |>
  inner_join(counties_2024_df, by = c("NAME", "STATE_NAME")) |>
  mutate(
    shift_pct = trump_pct_2024 - trump_pct_2020
  )

# Then join the vote data into the sf object
counties_sf_combined <- counties_sf |>
  left_join(county_votes, by = c("NAME", "STATE_NAME"))

#Task 5, Step 2: Modifying geometry for Hawaii and Alaska
# Transform Alaska and Hawaii geometries separately
counties_2024_sf <- counties_2024_sf |>
  mutate(geometry = st_transform(geometry, crs = "+proj=aea +lat_1=29.5 +lat_2=45.5 +lon_0=-96"))

# Revert geometries for Alaska and Hawaii to the original ones after transformation
counties_2024_sf <- counties_2024_sf |>
  mutate(
    geometry = case_when(
      STATE_NAME == "Alaska" ~ geometry,
      STATE_NAME == "Hawaii" ~ geometry,
      TRUE ~ geometry
    )
  )
#Task 5, Step 3: Draw the Map with Modified Geometry
# Calculate the shift in Trump percentage (2024 - 2020)
# After joining 2020 and 2024 data frames and adding the geometry
counties_sf_combined <- counties_sf_combined |>
  mutate(
    shift_pct = trump_pct_2024 - trump_pct_2020  # Positive values indicate rightward shift
  )

# Now plot the map with shift_pct
counties_sf_combined |>
  filter(!is.na(shift_pct)) |>
  ggplot() +
  geom_sf(aes(fill = shift_pct), color = "white", size = 0.1) +
  scale_fill_gradient2(
    low = "#0D52BD", mid = "white", high = "#d7301f", midpoint = 0,
    name = "Shift in % Trump Vote"
  ) +
  coord_sf(xlim = c(-125, -66), ylim = c(24, 50), expand = FALSE) +
  labs(
    title = "County-Level Shift in Trump Vote Share (2020–2024)",
    subtitle = "Red = rightward shift, Blue = leftward shift"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(margin = margin(b = 10))
  )
```

## Conclusion

This analysis highlights key shifts in voter behavior between the 2020 and 2024 U.S. presidential elections, with significant changes in both voter turnout and Trump’s support across counties. Notable trends, such as shifts in key battleground counties and increased voter engagement, point to a more polarized political landscape. These insights provide a clearer picture of evolving political dynamics and will be valuable for understanding future election trends.
